# Taking a look at covariate values across negative and positive points

We want the negative sampled points to represent the positive points well and
have similar distributions across all the covariates. 

First, we need to make sure our data is loaded.  

```{r}
library(terra)

data_path <- system.file("elev_scottsburg", package = "TerrainWorksUtils")
elev <- rast(file.path(data_path, "elev_scottsburg.flt"))
grad <- rast(file.path(data_path, "grad_40.tif"))
plan <- rast(file.path(data_path, "plan_40.tif"))
prof <- rast(file.path(data_path, "prof_40.tif"))
dev <- rast(file.path(data_path, "dev_40.tif"))
#pca <- rast(file.path(data_path, "pca_15m_48hr.tif"))

vars_raster <- c(
  grad,
  plan,
  prof,
  dev
)

initiation_points <- terra::vect(file.path(system.file("examples", package = "TerrainWorksUtils"),"initiation_points.shp"))
```

Next, we need to establish our negative regions and sample negative points. 

```{r}

training_data <- create_training_data_with_buffer(initiation_points, 
                                                  vars_raster, 
                                                  pos_buffer = 15, 
                                                  neg_buffer = 150)
```

And generate histograms for the distributions between different points. 

```{r}


sample_point_plots(training_data, 
                   c("dev_40", "grad_40", "plan_40", "prof_40"),
                   plot_type = "density")
```

We also want to look at the distributions from the entire positive and negative 
regions. We will generate polygons of the positive space and negative space, 
then extract values and plot it. 

```{r}

pos_region = pos_region <- terra::buffer(initiation_points,
                              width = 15 * 2)
pos_region <- terra::aggregate(pos_region)

neg_region <- make_neg_region(initiation_points,
                              vars_raster,
                              15,
                              150,
                              return_raster = FALSE)

pos_region_data <- extract_values(vars_raster, 
                                  pos_region, 
                                  extraction_method = "all")
pos_region_data$class <- "positive"

neg_region_data <- extract_values(vars_raster, 
                                  neg_region, 
                                  extraction_method = "all")
neg_region_data$class <- "negative"

```

And now graph these. 

```{r}
sample_point_plots(rbind(pos_region_data, neg_region_data), 
                   c("dev_40", "grad_40", "plan_40", "prof_40"), 
                   plot_type = "density")
```

