---
title: "DEMutilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DEMutilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TerrainWorksUtils)
```

This package includes some methods from the [DEM utilities toolbox](https://github.com/TerrainWorks-Seattle/ForestedWetlands/tree/master/DEMutilities). You can calculate slope metrics which can be used as model inputs. These methods are currently only supported on Windows. 

Available metrics include: 

 * `"grad"`: Surface gradient. 
 * `"plan"`: Plan curvature. Curvature along contour line (negative indicates convergent topography, positive indicates divergent topography)
 * `"prof"`: Profile curvature. A measure of changes in slope along the fall line (direction of maximum slope). Negative values indicate decreasing slope; positive values indicate increasing slope. Saturated return flow, where shallow groundwater tends to exfiltrate onto the surface, tends to occur in areas of negative profile curvature.
 * `"bcon"`
 * `"dev"`: Elevation deviation: Local deviation from mean elevation, an indicator of locally high or low topography. Defined as (z-zmean)/sd, where z is elevation at a point (DEM cell), zmean is mean elevation over a circle with diameter equal to the specified length scale, and sd is standard deviation of elevation within that circle.
 * `"twi"`: Topographic Wetness Index. 
 * `"pca"`: Partial contributing area. Not currently supported. 

Surface metrics can be calculated over a scale which is greater than the 
resolution of the DEM, which can add some smoothing to the output. The 
default is a 15-meter scale. 

Let's say you want to calculate gradient, plan curvature, and profile curvature
to use as inputs for a landslide susceptibility model. The example DEM 
which comes with this package is at a 20-meter scale, so we will calculate the 
metrics over a 40-meter scale. 

Let's create a temporary output directory to put the files. In your workflow, 
you should save them somewhere that makes sense, like the same directory as 
the DEM. 

```{r calculate-metrics}
DEM_path <- system.file("examples/elev_scottsburg.tif", 
                        package = "TerrainWorksUtils")
output_dir <- tempdir()
metrics <- c("plan", "grad", "prof")
calculate_surface_metrics(metrics = metrics, 
                          DEM_path = DEM_path, 
                          output_dir = output_dir, 
                          length_scale = 40)
```

This will have written rasters to the output directory for plan curvature, 
gradient, and profile curvature based on the DEM. They are written both as
`.flt` files and as `.tif` files. 

```{r list-files}
list.files(output_dir)
```

The input file used to run the executable is also saved to the output directory: 
`input_makeGrids.txt`. You can examine this file to help with debugging. 

