---
title: "Success Rate Curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Success Rate Curve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TerrainWorksUtils)
```

TerrainWorks has developed a few ways of evaluating models 

## Training data, etc. 

## The Success Rate Curve

We show how to generate the success rate curve using the function included in the package. 
Show what the output looks like. 
Show a graph of the success rate curve. 

Include an explanation and some references to the success rate curve in the literature. 


## The Suislaw Basin

[A map showing the basin and its predicted probabilities with one model, then another model.]

## The Prediction Rate Curve

Using the function to generate a prediction rate curve with predicted probabilities for the entire region. 

## Combining curves

Treat the Suislaw basin as two separate curves for this example. 

## Including it in the workflow 
