---
output:
  html_document: default
  pdf_document: default
---
# Generate Elevation Derivatives

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(TerrainWorksUtils)
```

You can use the function elev_deriv to calculate a variety of elevation-derived metrics
that can be used as model inputs. Elev_deriv calls a compiled Fortran program called makegrids.
The code is currently compiled for windows only. 

The following metrics are used for estimating landslide susceptibility: 

 * Surface gradient. 
 * Surface normal and tangential curvature. These can be summed and divided by 2 to get mean curvature.
 * Partial contributing area. This is the upslope contributing area to a DEM cell for a storm
   of specified duration. You need to specify the saturated hydraulic conductivity as well.

All surface metrics can be calculated over a length scale which is greater than the 
resolution of the DEM. This provides smoothing to the output. The default is 15 meters. 

The example DEM which comes with this package is at a 2-meter horizontal resolution and we will 
calculate the elevation derivatives over the default 15-meter scale. 

Let's create a temporary output directory to put the files. For this example, we create a temporary directory on the c drive. In your workflow, you should save them somewhere that makes sense, like the same directory as the DEM. Note that producing these rasters is computationally intensive and can take some time, depending on how large your DEM is. 

```{r calculate-metrics}
getwd()
DEM <- paste0(system.file("/inst/data/", package = "TerrainWorksUtils"), "/elev_BigCrSmith.flt")
output_dir <- "c:/tempDir/"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

length_scale = 15.0
rasters <- c(paste0("GRADIENT,", output_dir, "grad_15"),
             paste0("NORMAL SLOPE CURVATURE,", output_dir, "norm_15"),
             paste0("TANGENTIAL CURVATURE,", output_dir, "tan_15"))

elev_deriv(rasters = rasters,
           length_scale = length_scale,
           dem = DEM,
           scratch_dir = output_dir)

k = 1.0 # saturated hydraulic conductivity in meters per hour
d = 5.0 # storm duration in hours

contributing_area(raster = paste0(output_dir, "pca_k1_d5.flt"),
                  dem = DEM, 
                  length_scale = length_scale, 
                  k = k, 
                  d = d, 
                  scratch_dir = output_dir)

d = 10.0
contributing_area(raster = paste0(output_dir, "pca_k1_d10.flt"),
                  dem = DEM, 
                  length_scale = length_scale, 
                  k = 1, 
                  d = d, 
                  scratch_dir = output_dir)

d = 20.0
contributing_area(raster = paste0(output_dir, "pca_k1_d20.flt"),
                  dem = DEM, 
                  length_scale = length_scale, 
                  k = k, 
                  d = d, 
                  scratch_dir = output_dir)
```

This will have written rasters to the output directory for gradient, slope-normal and tangent curvature, and partial contributing area for three storm durations, all based on the input DEM. They are written as
`.flt` files. 

```{r list-files}
list.files(output_dir)
```

The input files used to run the executable is also saved to the output directory: 
`makeGrids_input.txt` and `partial_input.txt`. You can examine this file to help with debugging. 