# Training data sampling strategy.

```{r}
library(terra)
```

This example uses the elevation derivatives created with "Generate Elevation Derivatives" Markdown document. We'll use the same DEM and look for the created rasters in the c:/tempDir directory created with that document.

```{r}

DEM <- paste0(system.file("/inst/data/", package = "TerrainWorksUtils"), "/elev_BigCrSmith.flt")
elev <- rast(DEM)
output_dir <- "c:/tempDir/"
grad <- rast(paste0(output_dir, "grad_15.flt"))
norm <- rast(paste0(output_dir, "norm_15.flt"))
tan <- rast(paste0(output_dir, "tan_15.flt"))
pca5 <- rast(paste0(output_dir, "pca_k1_d5.flt"))
pca10 <- rast(paste0(output_dir, "pca_k1_d10.flt"))
pca20 <- rast(paste0(output_dir, "pca_k1_d20.flt"))

#calculate mean slopes
mean <- (norm + tan) / 2
names(mean) <- "mean_15"

vars_raster <- c(
  grad,
  mean,
  pca5,
  pca10,
  pca20
)

initiation_points <- vect(paste0(system.file("/inst/data/", package = "TerrainWorksUtils"), "/initiation_points.shp"))

```

Make a shaded relief image using the elevation raster and overlay the elevation and initiation points.

```{r}
slope <- terrain(elev, "slope", unit="radians")
aspect <- terrain(elev, "aspect", unit="radians")
elev_shade <- shade(slope, aspect, 45, 315)
plot(elev_shade, col=grey(1:100/100))
plot(elev, col = topo.colors(255), main = "topo.colors()", alpha=0.5, add=TRUE)
plot(initiation_points, add=TRUE)
```

Create buffers around the initiation points to build the negative (no landslides) region. 

We also build an analysis mask, which selects only the pixels where selected 
elevation derivatives are within the range that we see in initiation points.
This step would eliminate areas like stream beds that we can be confident are 
not prone to landslide intiation. 

```{r}
mask_rasters <- c("grad_15", "mean_15")
analysis_mask <- create_analysis_region_mask(raster = vars_raster, 
                                             points = initiation_points,
                                             mask_vars = mask_rasters)

neg_region <- make_neg_region(initiation_points, 
                vars_raster, 
                inner_buffer = 15., 
                outer_buffer = 250.,
                return_raster=TRUE)

final_mask <- analysis_mask * neg_region

plot(elev_shade, col=grey(1:100/100))
plot(final_mask, col=rgb(1,0,0,0.5), add=TRUE)
plot(initiation_points, add=TRUE)

```

Sample negative points. Here we set a random seed to ensure that the results 
could be replicated. 

```{r}
training_points <- sample_negative_points(
    positive_points = initiation_points,
    analysis_region = analysis_mask,
    buffer = TRUE,
    buffer_radius = 11,
    negative_proportion = 1., 
    rseed = 2
  )
```

Plot the study area with positive and negative sampled points. Positive points
are shown in black, while the negative sampled points are shown in red. 

```{r}
plot(elev_shade, col=grey(1:255/255))
plot(training_points, col = ifelse(training_points$class == "positive", 
                                    "black", "red"), add=TRUE)
```

Next, extract the values and put them into a training data frame for model input.

```{r}
training_data <- extract_values(
    raster = vars_raster,
    points = training_points,
    extraction_method = "centroid",
    extraction_layer = NULL,
    xy = FALSE
  )
```

