# Training data sampling strategy.

```{r}
library(terra)
```

Load the data into R. Right now, I have higher resolution data stored in a 
folder in sandbox/data. These files can be recalculated from the 
00_generate_elev_derivatives script. 

Replace the data_path to use data. 

```{r}
data_path <- "\\\\Mac/Home/Documents/terrainworks/code/sandbox/data/DeanCr/"

elev <- rast(file.path(data_path, "elev_scottsburg.flt"))
grad <- rast(file.path(data_path, "grad_15.flt"))
plan <- rast(file.path(data_path, "plan_15.flt"))
prof <- rast(file.path(data_path, "prof_15.flt"))
norm <- rast(file.path(data_path, "norm_15.flt"))
tan <- rast(file.path(data_path, "tan_15.flt"))
pca <- rast(file.path(data_path, "pca_k1_d5.flt"))

#calculate mean slopes
mean <- (norm + tan) / 2
names(mean) <- "mean_15"

vars_raster <- c(
  grad,
  plan,
  prof,
  norm, 
  tan, 
  mean,
  pca
)

initiation_points <- vect(file.path(data_path, "initiation_points.shp"))

```

Plot elevation + initiation points.

```{r}
plot(elev)
points(initiation_points)
```

Create buffers around the initiation points to build our negative region. 

We also build an analysis mask, which selects only the pixels where selected 
elevation derivatives are within the range that we see in initiation points.
This step would eliminate areas like stream beds that we can be confident are 
not prone to landslide intiation. 

```{r}
neg_region <- make_neg_region(initiation_points, 
                vars_raster, 
                inner_buffer = 11, 
                outer_buffer = 150)

analysis_mask <- create_analysis_region_mask(raster = vars_raster, 
                                             points = initiation_points)

```

Plot the study region with initiation points and the negative region.

```{r}

plot(elev)
points(initiation_points)
plot(terra::mask(neg_region, analysis_mask), add = TRUE, col = rgb(.2, .2, .2, alpha = .5))

```

Sample negative points. Here we set a random seed to ensure that the results 
could be replicated. 

```{r}
training_points <- sample_negative_points(
    positive_points = initiation_points,
    analysis_region = analysis_mask,
    buffer = TRUE,
    buffer_radius = 11,
    negative_proportion = 1., 
    rseed = 2
  )
```

Plot the study area with positive and negative sampled points. Positive points
are shown in black, while the negative sampled points are shown in red. 

```{r}
plot(elev)
plot(analysis_mask, add = TRUE)
points(training_points, col = ifelse(training_points$class == "positive", 
                                    "black", 
                                    "red"))
```

Next, extract the values and put them into a training data frame for model input.

```{r}
training_data <- extract_values(
    raster = vars_raster,
    points = training_points,
    extraction_method = "centroid",
    extraction_layer = NULL,
    xy = FALSE
  )
```

