# Analyze the training data produced to assess the negative sampling strategy. 

Here we investigate how the sampling strategy works to create useful input data for model generation. 

First, load the elevation derivatives. These are calculate in (00) generate_elev_derivatives. 

Replace data_path as needed.

```{r}
data_path <- "\\\\Mac/Home/Documents/terrainworks/code/sandbox/data/DeanCr"
# data_path <- "\\\\Mac/Home/Documents/terrainworks/code/sandbox/data/scottsburg"

elev <- terra::rast(file.path(data_path, "elev_deancr.flt"))
grad <- rast(file.path(data_path, "deriv/grad_15.flt"))
plan <- rast(file.path(data_path, "deriv/plan_15.flt"))
prof <- rast(file.path(data_path, "deriv/prof_15.flt"))
norm <- rast(file.path(data_path, "deriv/norm_15.flt"))
tan <- rast(file.path(data_path, "deriv/tan_15.flt"))
pca_5 <- rast(file.path(data_path, "deriv/pca_k1_d5.flt"))
pca_10 <- rast(file.path(data_path, "deriv/pca_k1_d10.flt"))
pca_20 <- rast(file.path(data_path, "deriv/pca_k1_d20.flt"))

#calculate mean slopes
mean <- (norm + tan) / 2
names(mean) <- "mean_15"

vars_raster <- c(
  grad,
  plan,
  prof,
  norm, 
  tan, 
  mean,
  pca_5, 
  pca_10,
  pca_20
)

initiation_points <- vect(file.path(data_path, "DeanCr_Initiation_Points.shp"))
```

Generate the negative region and analysis mask for visualization. 

```{r}
neg_region <- make_neg_region(initiation_points, 
                vars_raster, 
                inner_buffer = 11, 
                outer_buffer = 200)
analysis_mask <- create_analysis_region_mask(raster = vars_raster, 
                                             points = initiation_points)
```

```{r}
plot(elev)
points(initiation_points)
plot(terra::mask(neg_region, analysis_mask), add = TRUE, col = rgb(.2, .2, .2, alpha = .5))
```


Now, with the help of random seeding, we want to repeat the training data generation a couple of times, and compare the varying distributions. 

```{r}

seeds = c(1, 50, 32, 56, 3)

training_data <- create_training_data_with_buffer(
  positive_points = initiation_points,
  predictors_raster = vars_raster,
  pos_buffer = 11,
  neg_buffer = 150,
  analysis_region_mask = NULL,
  negative_proportion = 1,
  extraction_method = "centroid",
  extraction_layer = NULL,
  rseed = seeds[2])

```

Plot the distributions of elevation derivatives for positive points vs. negative
sampled points. The positive distribution is shown in black and the negative 
points are shown in red. 

```{r}
plot_metric_distribution <- function(training_data,
                                     parameters = NULL,
                                     plot_type = "hist") {

  if (!(plot_type %in% c("hist", "den"))) {
    stop("Plot type must be either \"hist\" (histogram) or \"den\" (density)")
  }

  # separate out the positive and negative data
  positive_data <- training_data[which(training_data$class == "positive"), ]
  negative_data <- training_data[which(training_data$class == "negative"), ]

  if (is.null(parameters)) {
    parameters = names(training_data)
  }

  # define visual aspects of the plot
  num_parameters = length(parameters)
  par(mfrow = c(ceiling(num_parameters / 3), 3), mar = c(2, 2, 1, 1),
      oma = c(2,1,1,1))

  blk <- rgb(0, 0, 0, max = 255, alpha = 80)
  red <- rgb(227, 60, 57, max = 255, alpha = 80)

  if (plot_type == "hist") {
    for (p in parameters) {
      # use consistent buckets for the histograms
      buckets = seq(min(training_data[, p], na.rm = TRUE),
                    max(training_data[, p], na.rm = TRUE),
                    length.out = 20)

      if (plot_type == "hist") {
        h_pos <- hist(positive_data[, p], breaks = buckets, plot = FALSE)
        h_neg <- hist(negative_data[, p], breaks = buckets, plot = FALSE)
        plot(h_neg, xlab = p,
             main = paste0("Distribution of ", p),
             col = red,
             ylim = c(0, max(c(h_pos$counts, h_neg$counts))))
        plot(h_pos, col = blk,
             add = TRUE)
      }
    }
    } else {
      for (p in parameters) {
        d_pos <- density(positive_data[, p], n = 1024)
        d_neg <- density(negative_data[, p], n = 1024)
        plot(d_pos,
             main = paste0("Distribution of ", p),
             xlab = p,
             col = blk,
             ylim = c(0, max(c(d_pos$y, d_neg$y))))
        polygon(d_pos, col = blk)
        polygon(d_neg, col = red)
      }
      # plot(1, type = "n", axes=FALSE, xlab="", ylab="")
      # plot_colors <- c(rgb(0, 0, 0, max = 255), rgb(227, 60, 57, max = 255))
      # legend(x = "top",inset = 0,
      #        legend = c("Positive points", "Negative points"), 
      #        col=plot_colors, lwd=5, cex=.5, horiz = TRUE)
      
      par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
      plot(0, 0, type = 'l', bty = 'n', xaxt = 'n', yaxt = 'n')
      legend('bottom',legend = c("Positive points", "Negative points"), 
             col = c(rgb(0, 0, 0, max = 255), rgb(227, 60, 57, max = 255)), 
             lwd = 5, xpd = TRUE, horiz = TRUE, cex = 1, seg.len=1, bty = 'n')
  }
}

cols = c("grad_15", "plan_15", "prof_15", "norm_15", "tan_15", "mean_15", "pca_k1_d5", "pca_k1_d10", "pca_k1_d20")
plot_metric_distribution(training_data,
                         parameters = cols,
                         plot_type = 'den')

```

Extract values for the entire negative region. 

```{r}

neg_reg <- terra::extract(vars_raster, as.polygons(neg_region))

```

Plot the distribution of elevation derivatives for the entire negative region 
vs sampled points within the negative region. 

```{r}
# define visual aspects of the plot
num_parameters = length(cols)
par(mfrow = c(ceiling(num_parameters / 3), 3), mar = c(2, 2, 1, 1),
      oma = c(2,1,1,1))
blk <- rgb(0, 0, 0, max = 255, alpha = 80)
red <- rgb(227, 60, 57, max = 255, alpha = 80)

column = "plan_15"

for (column in cols) {
  
    d_reg <- density(neg_reg[, column], n = 1024, na.rm = TRUE)
    d_pts <- density(training_data[which(training_data$class == "negative"), column], n = 1024) 
    
    suppressWarnings(res <- ks.test(training_data[which(training_data$class == "negative"), column], 
                   neg_reg[, column]))

    
    plot(d_reg,
         main = paste0("Distribution of ", column, "(p = ", format(round(res$p.value, 3), nsmall = 3), ")"),
         xlab = column,
         col = blk,
         ylim = c(0, max(c(d_reg$y, d_pts$y))),
         xlim = c(min(d_pts$x), quantile(d_pts$x, probs = c(.95))))
         # xlim = c(min(d_))
    polygon(d_reg, col = blk)
    polygon(d_pts, col = red)
      
}

par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
plot(0, 0, type = 'l', bty = 'n', xaxt = 'n', yaxt = 'n')
legend('bottom',legend = c("Negative region", "Sampled negative points"), 
       col = c(rgb(0, 0, 0, max = 255), rgb(227, 60, 57, max = 255)), 
       lwd = 5, xpd = TRUE, horiz = TRUE, cex = 1, seg.len=1, bty = 'n')

```


