---
title: "Principal component analysis on landslide training data"
author: Julia Lober
output:
  html_notebook:
    df_print: paged
  html_document: 
    toc: true
params:
  data_dir:
    label: Data Directory
    value: //Mac/Home/Documents/terrainworks/code/sandbox/data/
    input: text
  training_data:
    label: Training data .Rdata file
    value:  /sample_umpqua.Rdata
    input: text
editor_options:
  chunk_output_type: inline
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

# Principal component analysis
A principal component analysis (PCA) allows us to reduce the dimensionality of data and visualize the variation in a dataset. It is usually an exploratory analysis rather than a key part of a quantitative analysis. Here, we do it to see where different covariates might be correlated and how the initiation points vs. non-inititation points are reflected in the overall variability. 

```{r}

load(paste0(params$data_dir, params$training_data))

pc <- prcomp(sample_pnts[,2:7],
             center = TRUE,
             scale. = TRUE)

plot(pc$x[sample_pnts$class == "neg",], pch = 19)
points(pc$x[sample_pnts$class == "pos",], col = "red", pch = 19)

legend("topright", 
       legend = c("Negative points", "Initiation points"), 
       pch = 19, 
       col = c("black", "red"),
       bty= "n")


```

